# Reproducible Research: Peer Assignment 1


```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(plyr)
```

## Loading and preprocessing the data

###### 1. Load the data 
```{r, results='markup', warning=TRUE, message=TRUE}
activity <- read.csv("activity.csv")
```

###### 2. Process/transform the data into a format suitable for the analysis
```{r}
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")

clean <- activity[!is.na(activity$steps),]
```


## What is mean total number of steps taken per day?

###### 1. Calculate the total number of steps taken per day.
```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum )
colnames(sumTable)<- c("Date", "Steps")
```

###### 2. Make a histogram of the total number of steps taken each day
```{r}
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

###### 3. Calculate and report the mean and median total number of steps taken per day
```{r}
## Mean of Steps
stepsByDayMean <- as.integer(mean(sumTable$Steps))

## Median of Steps
stepsByDayMedian <- as.integer(median(sumTable$Steps))
```
* The average number of steps taken each day was `r stepsByDayMean` steps.
* The median number of steps taken each day was `r stepsByDayMedian` steps.

## What is the average daily activity pattern?

###### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
library(plyr)
library(ggplot2)
##pulling data without nas
clean <- activity[!is.na(activity$steps),]

##create average number of steps per interval
intervalTable <- ddply(clean, .(interval), summarize, Avg = mean(steps))

##Create line plot of average number of steps per interval
p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")

```

###### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
##Maximum steps by interval
maxSteps <- max(intervalTable$Avg)
A_maxSteps <- as.integer(max(intervalTable$Avg))
##Which interval contains the maximum average number of steps
time <- intervalTable[intervalTable$Avg==maxSteps,1]
```
* The maximum number of steps for a 5-minute interval was `r A_maxSteps` steps.
* The `r time`th interval has the maximum number of steps.

## Imputing missing values

###### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
##Number of NAs in original data set
count <- nrow(activity[is.na(activity$steps),])
```

* The total number of rows with steps = 'NA' is `r count`

###### 2. Devise a strategy for filling in all of the missing values in the dataset. 

```{r}
## Create the average number of steps per weekday and interval
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))

## Create dataset with all NAs for substitution
nadata<- activity[is.na(activity$steps),]
## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))

## Reorder the new substituded data in the same format as clean data set
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

##Merge the NA averages and non NA data together
mergeData <- rbind(clean, newdata2)

```
###### 3. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
##Create sum of steps per date to compare with step 1
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum )
colnames(sumTable2)<- c("Date", "Steps")

## Mean of Steps with NA data taken care of
A_Mean <- as.integer(mean(sumTable2$Steps))

## Difference from the original mean
diff_Mean <- A_Mean - stepsByDayMean

## Median of Steps with NA data taken care of
A_Median <- as.integer(median(sumTable2$Steps))

## Difference from the original median
diff_Median <- A_Median - stepsByDayMedian

## Creating the histogram of total steps per day, categorized by data set to show impact
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Red")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Green", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("Red", "Green") )

```


* The new mean of the imputed data is `r A_Mean` steps compared to the old mean of `r stepsByDayMean` steps. The difference is `r diff_Mean` steps.

* The new median of the imputed data is `r A_Median` steps compared to the old median of `r stepsByDayMedian`. The difference is `r diff_Median` steps.


## Are there differences in activity patterns between weekdays and weekends?

###### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
## Create new category based on the days of the week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```
###### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
library(lattice) 
## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```

